---
output:
  rmarkdown::github_document
bibliography: "README-refs.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# R/`haldensify`

[![Travis-CI Build Status](https://travis-ci.org/nhejazi/haldensify.svg?branch=master)](https://travis-ci.org/nhejazi/haldensify)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/nhejazi/haldensify?branch=master&svg=true)](https://ci.appveyor.com/project/nhejazi/haldensify)
[![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/haldensify/master.svg)](https://codecov.io/github/nhejazi/haldensify?branch=master)
[![CRAN](http://www.r-pkg.org/badges/version/survsl3)](http://www.r-pkg.org/pkg/survsl3)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/survsl3)](https://CRAN.R-project.org/package=survsl3)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Nonparametric Conditional Density Estimation with the Highly Adaptive Lasso

__Author:__ [Nima Hejazi](https://nimahejazi.org)

---

## What's `haldensify`?

The `haldensify` R package is designed to provide facilities for nonparametric
conditional density estimation based on the procedure proposed by
@diaz2011super. The core procedure involves recovering conditional density
estimates by performing pooled hazards regressions on a long format data set
built from the input data. For the time being, `haldensify` is a minimal
implementation of this density estimation strategy, though future generalization
of the core routines may be possible.

---

## Installation

Install the most recent _stable release_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-master-installation, eval = FALSE}
devtools::install_github("nhejazi/haldensify", build_vignettes = FALSE)
```

---

## Example

A simple example illustrates how `haldensify` may be used to construct
conditional density estimates:

```{r example, message=FALSE, warning=FALSE}
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/nhejazi/haldensify/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/nhejazi/haldensify/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `haldensify` R package, please cite the following:

        @manual{hejazi2019haldensify,
          author = {Hejazi, Nima S},
          title = {haldensify: Nonparametric conditional density estimation
            with the highly adaptive lasso in {R}},
          year  = {2019},
          url = {https://github.com/nhejazi/haldensify},
          note = {R package version 0.0.1}
        }

---

## Related

* [R/`condensier`](https://github.com/osofr/condensier) - An  independent
  implementation of the same core methodology, though more general in its making
  allowance for arbitrary selection of regression functions and a greater
  variety of hazard regression strategies.

---

## License

&copy; 2019 [Nima S. Hejazi](https://nimahejazi.org)

The contents of this repository are distributed under the MIT license. See below
for details:
```
MIT License

Copyright (c) 2019 Nima S. Hejazi

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.
```

---

## References

