---
output:
  rmarkdown::github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# R/`haldensify`

[![Travis-CI Build Status](https://travis-ci.org/nhejazi/haldensify.svg?branch=master)](https://travis-ci.org/nhejazi/haldensify)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/nhejazi/haldensify?branch=master&svg=true)](https://ci.appveyor.com/project/nhejazi/haldensify)
[![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/haldensify/master.svg)](https://codecov.io/github/nhejazi/haldensify?branch=master)
[![CRAN](http://www.r-pkg.org/badges/version/haldensify)](http://www.r-pkg.org/pkg/haldensify)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/haldensify)](https://CRAN.R-project.org/package=haldensify)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3698329.svg)](https://doi.org/10.5281/zenodo.3698329)

> Highly Adaptive Lasso Conditional Density Estimation

__Authors:__ [Nima Hejazi](https://nimahejazi.org), [David
Benkeser](https://www.sph.emory.edu/faculty/profile/#!dbenkes), and [Mark
van der Laan](https://vanderlaan-lab.org/about/)

---

## What's `haldensify`?

The `haldensify` R package is designed to provide facilities for nonparametric
conditional density estimation based on a flexible procedure proposed initially
by @diaz2011super. The core of the implemented methodology involves recovering
conditional density estimates by performing pooled hazards regressions so as to
assess the conditional hazard that an observed value falls in a given bin over
the (conditional) support of the variable of interest. Such conditional density
estimates are useful, for example, in causal inference problems in which the
_generalized propensity score_ (for continuous-valued exposures) must be
estimated [@diaz2012population; @diaz2018stochastic; @diaz2020causal].
`haldensify` implements this condtional density estimation strategy
specifically for use only with the highly adaptive lasso [@benkeser2016highly;
@vdl2017generally; @vdl2018highly; @coyle2019hal9001].

---

## Installation

For standard use, we recommend installing the package from
[CRAN](https://CRAN.R-project.org/package=haldensify) via

```{r cran-installation, eval = FALSE}
install.packages("haldensify")
```

To contribute, install the _development version_ of `haldensify` from GitHub
via [`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval = FALSE}
remotes::install_github("nhejazi/haldensify")
```

---

## Example

A simple example illustrates how `haldensify` may be used to construct
conditional density estimates:

```{r example, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(haldensify)
set.seed(76924)

# simulate data: W ~ U[-4, 4] and A|W ~ N(mu = W, sd = 0.5)
n_train <- 100
w <- runif(n_train, -4, 4)
a <- rnorm(n_train, w, 0.5)

# learn relationship A|W using HAL-based density estimation procedure
mod_haldensify <- haldensify(
  A = a, W = w,
  n_bins = c(10, 20),
  grid_type = c("equal_range", "equal_mass"),
  lambda_seq = exp(seq(-1, -10, length = 300))
)

# predictions to recover conditional density of A|W
new_a <- seq(-4, 4, by = 0.1)
new_w_neg <- rep(-2, length(new_a))
new_w_zero <- rep(0, length(new_a))
new_w_pos <- rep(2, length(new_a))
new_dat <- as.data.table(list(a = new_a, w_neg = new_w_neg,
                              w_null = new_w_zero, w_pos = new_w_pos))
new_dat$pred_w_neg <- predict(mod_haldensify,
                              new_A = new_dat$a, new_W = new_dat$w_neg)
new_dat$pred_w_null <- predict(mod_haldensify,
                               new_A = new_dat$a, new_W = new_dat$w_null)
new_dat$pred_w_pos <- predict(mod_haldensify,
                              new_A = new_dat$a, new_W = new_dat$w_pos)

# visualize results
p <- new_dat %>%
  melt(id = c("a"), measure.vars =
       c("pred_w_pos", "pred_w_null", "pred_w_neg")) %>%
  ggplot(aes(x = a, y = value, colour = variable)) +
  geom_point() +
  geom_line() +
  stat_function(fun = dnorm, args = list(mean = -2, sd = 0.5),
                colour = "blue", linetype = "dashed") +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 0.5),
                colour = "darkgreen", linetype = "dashed") +
  stat_function(fun = dnorm, args = list(mean = 2, sd = 0.5),
                colour = "red", linetype = "dashed") +
  xlab("Observed value") +
  ylab("Predicted probability") +
  ggtitle("Conditional density p(A|W)") +
  theme_bw() +
  theme(legend.position = "none")
p %>% print()
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/nhejazi/haldensify/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/nhejazi/haldensify/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `haldensify` R package, please cite the following:

        @manual{hejazi2020haldensify,
          author = {Hejazi, Nima S and Benkeser, David C and {van der Laan},
            Mark J},
          title = {{haldensify}: Highly adaptive lasso conditional density
            estimation },
          year  = {2020},
          howpublished = {\url{https://github.com/nhejazi/haldensify}},
          note = {{R} package version 0.0.5},
          url = {https://doi.org/10.5281/zenodo.3698329},
          doi = {10.5281/zenodo.3698329}
        }

---

## Related

* [R/`hal9001`](https://github.com/tlverse/hal9001) -- The highly adaptive
  lasso estimator used internally to constructed conditional density estimates.
* [R/`condensier`](https://github.com/osofr/condensier) -- An  independent
  implementation of some of the same core methodology, allowing for arbitrary
  selection of regression functions and a greater variety of hazard regression
  strategies.

---

## License

&copy; 2019-2020 [Nima S. Hejazi](https://nimahejazi.org)

The contents of this repository are distributed under the MIT license. See below
for details:
```
MIT License

Copyright (c) 2019-2020 Nima S. Hejazi

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.
```

---

## References

